---
title: "T1/T2D"
author: "J.Knupp"
date: "2023-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary libraries
```{r libraries}
#Set working directory
setwd("~/T1T2D")
#load necessary packages
#haven allows you to read in stata datasets (after stat version 12 - v5-12 use library(foreign)) into R
library(haven)
library(dplyr)
library(ggplot2)
library(psych)
library(epiDisplay)
#for running resticted cubic splines
library(rms)
library(Hmisc)
#for providing aic for multple models
library(AICcmodavg)
library(jtools)

```

## 1. Dataset
### 1.1 Read in dataset
Read in dataset and filter relevant variables into new data.table. These are:
-*outcome*: insulinRequire
-*predictor variables*:
-AgeatDiagnosis
-bmi
-sexMale
-GADPositive975
-IA2Positive975
-z1w_score
-HBA1CDescriptive
```{r dataset}
#read in dataset
clinicalfeatures <- read_dta("clinicalFeaturesModelV2.dta")
#filter out relevant variables
cf <- subset(clinicalfeatures, select = c("AgeatDiagnosis", "insulinRequire", "bmi", "GADPositive975", "IA2Positive975", "sexMale", "z1w_score", "HBA1CDescriptive"))
```

### 1.2 Distributions of continuous traits
#### 1.2.1 Age at diagnosis (aad)
```{r aad distribution}
#numerial distribution
summary(cf$AgeatDiagnosis)
#visualising bmi distribution
aad_box <- ggplot(cf, aes(AgeatDiagnosis)) +
  geom_boxplot() +
  labs(title="Box and Whisker plot of Age at Diagnosis", 
       font="bold")
aad_box
#produce density histogram
ggplot(cf, aes(AgeatDiagnosis)) + 
  geom_histogram(aes(y=..density..),fill="bisque",color="white",alpha=0.7, binwidth = 2) + 
  geom_density() +
  geom_rug() +
  labs(x='Age at Diagnosis') +
  theme_minimal()
```
No outliers present for aad

#### 1.2.2 bmi
```{r bmi distribution}
#numerical distribution
summary(cf$bmi)
#visualising bmi distribution
bmi_box <- ggplot(cf, aes(bmi)) +
  geom_boxplot() +
  labs(title="Box and Whisker plot of BMI", 
       font="bold")
bmi_box

#distribution measures for bmi
Q1_bmi <- quantile(cf$bmi, 0.25)
Q3_bmi <- quantile(cf$bmi, 0.75)
IQR_bmi <- Q3_bmi - Q1_bmi
upp_thresh_bmi <- Q3_bmi + 1.5*IQR_bmi
low_thresh_bmi <- Q1_bmi - 1.5*IQR_bmi

#identify bmi outliers
bmi_out <- cf %>%
  filter(bmi < low_thresh_bmi | bmi > upp_thresh_bmi)
head(bmi_out)

#produce density histogram
ggplot(cf, aes(bmi)) + 
  geom_histogram(aes(y=..density..),fill="bisque",color="white",alpha=0.7, binwidth = 2) + 
  geom_density() +
  geom_rug() +
  labs(x='bmi') +
  theme_minimal()
```
21 outliers for BMI

### 1.3 Relationship between continuous predictors
```{r correlation}
#scatterplot btwn aad and bmi
scat_aad_bmi <- ggplot(cf, aes(AgeatDiagnosis, bmi)) +
  geom_point()
scat_aad_bmi

#scatterplot matrix aad and bmi
scatmat_aad_bmi <- pairs(cf[c("AgeatDiagnosis", "bmi")])
scatmat_aad_bmi

#correlation coefficents
cor_aad_bmi <- cor(cf$bmi, cf$AgeatDiagnosis)
cor_aad_bmi
```

### 1.4 Outcome variable
#### 1.4.1 Frequency of binary outcome
```{r iR proportions}
#summarise outcomes (insulinRequire)
count_iR <- tab1(cf$insulinRequire, cum.percent = TRUE, main = "Insulin Required")
#proportions of whether require insulin (1) or not (0)
c <- cf %>%
  count(insulinRequire, name = "n") %>%
  mutate(prop = n/sum(n)) %>%
  filter(insulinRequire == "1") %>%
 mutate(odds = prop/(1 - prop),
    log_odds = log(prop/(1 - prop)))
c

```

#### 1.4.2 Distributions of continuous predictors based on binary outcome
```{r distributions of predictors based on outcome}
#bmi
#histogram based on outcome
ggplot(cf, aes(bmi)) +
  geom_histogram() +
  facet_wrap(vars(insulinRequire))
#summarise sorted by outcome
summ_iR_bmi <-  describeBy(cf$bmi, cf$insulinRequire)
#violin plot of bmi distrubution across whether require insulin(1) or not (0)
ggplot(cf, aes(x = bmi, y = insulinRequire)) +
 geom_violin() +
 ylab("Require Insulin")

#AgeatDiagnosis
#histogram based on outcome
ggplot(cf, aes(AgeatDiagnosis)) +
  geom_histogram() +
  facet_wrap(vars(insulinRequire))
#summarise sorted by outcome
summ_iR_aad <-  describeBy(cf$AgeatDiagnosis, cf$insulinRequire)
#violin plot of bmi distrubution across whether require insulin(1) or not (0)
ggplot(cf, aes(x = AgeatDiagnosis, y = insulinRequire)) +
 geom_violin() +
 ylab("Require Insulin")

#z1w_score
#histogram based on outcome
ggplot(cf, aes(z1w_score)) +
  geom_histogram() +
  facet_wrap(vars(insulinRequire))
#summarise sorted by outcome
summ_iR_z1w <-  describeBy(cf$z1w_score, cf$insulinRequire)

#HBA1CDescriptive
#histogram based on outcome
ggplot(cf, aes(HBA1CDescriptive)) +
  geom_histogram() +
  facet_wrap(vars(insulinRequire))
#summarise sorted by outcome
summ_iR_HBA1C <-  describeBy(cf$HBA1CDescriptive, cf$insulinRequire)

#scatterplots of bmi &aad by insulinRequire
ggplot(cf, aes(AgeatDiagnosis, bmi)) +
  geom_point() +
  facet_wrap(vars(insulinRequire))

#summarise aad, sorting by insulinRequire & sexMale
summ_iR_sex_aad <-  describeBy(cf$AgeatDiagnosis, list(cf$insulinRequire, cf$sexMale))

```

#### 1.4.3 Distributions of binary predictors based on outcome variable
```{r distrinbution binary predicotrs by outcome}
#summarise categorical vairables using two way tables
#include frequencies, relative frequencies, pearsons chi squared, expected frequencies
#sexMale and insulinRequire
#Q - how to I combine all metrics into one table, and how add row and column names?
#describe function from psych library - select columns, dplyr (dataset>select columns> describe)
sex_iR_tbl <- table(cf$sexMale, cf$insulinRequire)
#performs chiquared test and outputs as list
summary(sex_iR_tbl)
#barplot of two way tablechi
barplot(sex_iR_tbl)

#GADPositive975 and insulinRequire
GAD_iR_tbl <- table(cf$GADPositive975, cf$insulinRequire)
#performs chiquared test and outputs as list
summary(GAD_iR_tbl)
#barplot of two way tablechi
barplot(GAD_iR_tbl)

#IA2Positive975 and insulinRequire
IA2_iR_tbl <- table(cf$IA2Positive975, cf$insulinRequire)
#performs chiquared test and outputs as list
summary(IA2_iR_tbl)
#barplot of two way tablechi
barplot(IA2_iR_tbl)
```

## 2. Univariate logistic regression models
### 2.1 BMI
#### 2.1.1 Simple
```{r bmi simple log reg}
#logistic regression for bmi (continuous trait)
mdl_iR_vs_bmi <- glm(insulinRequire ~ bmi, data = cf, family = "binomial")
summary(mdl_iR_vs_bmi)
#perform prediction on same datasets - new variable containing the prediction
cf <- cf %>%
  mutate(pr_iR_vs_bmi = predict(mdl_iR_vs_bmi, cf, type = "response"))
#perform prediction on same datasets - new variable containing the prediction in log odds
cf <- cf %>%
  mutate(lo_iR_vs_bmi = predict(mdl_iR_vs_bmi, cf))
pr <- predict(mdl_iR_vs_bmi, cf, type = "response")
#convert to numeric
cf$bim <- as.numeric(cf$bmi)
cf$insulinRequire <- as.numeric(cf$insulinRequire)
#visualisation
#how to I include the abline?
ggplot(cf, aes(bmi, insulinRequire)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 20) +
  geom_smooth(method="glm", method.args = list(family = "binomial"))
#plotting - gives same plot essentially as prediction plot with ggplot above
effect_plot(mdl_iR_vs_bmi, pred = bmi, plot.points = TRUE,
            jitter = c(0.1, 0.05), point.alpha = 0.1) +
  ylab("Pr(insulinRequire = 1)")
```
##### 2.1.1.1 Binned vs log odds 
```{r bin bmi}
#data bining plot
#identify decile cut-offs
quantile(cf$bmi, probs = seq(0,1,0.1))
#set up cutoff values
breaks <- c(17.53, 24.73, 27.05, 28.97, 30.94, 32.75, 34.79, 36.68, 39.45, 43.998, 70.16)
#specify interval/bin labels
tags <- c("[17.53, 24.73)", "[24.73, 27.05)", "[27.05, 28.97)", "[28.97, 30.94)", "[30.94,32.75)", "[32.75, 34.79)", "[34.79, 36.68)", "[36.68, 39.45)", "[39.45, 43.998)", "[43.998, 70.16)")
#sort values into decile bins
v <- cf %>%
  dplyr::select(bmi, insulinRequire) #pick the variable 
vgroup <- as_tibble(v) %>% 
  mutate(tag = case_when(
    bmi >= 17.53 & bmi < 24.73 ~ tags[1],
    bmi >= 24.73 & bmi < 27.05 ~ tags[2],
    bmi >= 27.05 & bmi < 28.97 ~ tags[3],
    bmi >= 28.97 & bmi < 30.94 ~ tags[4],
    bmi >= 30.94 & bmi < 32.75 ~ tags[5],
    bmi >= 32.75 & bmi < 34.79 ~ tags[6],
    bmi >= 34.79 & bmi < 36.68 ~ tags[7],
    bmi >= 36.68 & bmi < 39.45 ~ tags[8],
    bmi >= 39.45 & bmi < 43.998 ~ tags[9],
    bmi > 43.998 ~ tags[10],
    ))
summary(vgroup)
vgroup <- vgroup %>%
  mutate(lo_iR_vs_bmi = predict(mdl_iR_vs_bmi, cf))
v_bin_log <- vgroup %>%
  group_by(tag) %>%
  summarise(median_score=median(lo_iR_vs_bmi))
ggplot(v_bin_log, aes(tag, median_score)) +
  geom_point() +
  labs(x = 'BMI decile bins', y = 'median log odds of T1D')

```

#### 2.1.2 Transformations

#### 2.1.3 Spline transformation
```{r bmi cubic splines}
#plot function requires variables to be in numeric format
#run with 3, 4 & 5 knots
#how do I get the summary stats to automatically be attached - error message when run in chunk
#rcspline.plot(as.numeric(cf$bmi), as.numeric(cf$insulinRequire), model="logistic", nk=3)
#rcspline.plot(as.numeric(cf$bmi), as.numeric(cf$insulinRequire), model="logistic", nk=4)
#rcspline.plot(as.numeric(cf$bmi), as.numeric(cf$insulinRequire), model="logistic", nk=5)

#run a non spline log regression, then one each with knots = 3, 4, 5 respectively
b0 <- lrm(insulinRequire ~ bmi, data = cf)
b1 <- lrm(insulinRequire ~ rcs(bmi, 3), data = cf)
b2 <- lrm(insulinRequire ~ rcs(bmi, 4), data = cf)
b3 <- lrm(insulinRequire ~ rcs(bmi, 5), data = cf)

dd <- datadist(cf)
options(datadist="dd")
summary(b0)
summary(b1)
summary(b2)
summary(b3)

#run anovas on each to see Wald statistics
anova(b0)
anova(b1)
anova(b2)
anova(b3)

#get aic
#why is this different to the plot AIC values for m1 etc.
bAIC <- c(b0 = AIC(b0), b1 = AIC(b1), b2 = AIC(b2), b3 = AIC(b3))
bAIC

```


### 2.2 Age at Diagnosis
#### 2.2.1 Simple
```{r aad log reg}
#logistic regression for AgeatDiagnosis (continuous trait)
mdl_iR_vs_aad <- glm(insulinRequire ~ AgeatDiagnosis, data = cf, family = "binomial")
summary(mdl_iR_vs_aad)
#perform prediction on same datasets - new variable containing the prediction
cf <- cf %>%
  mutate(pr_iR_vs_aad = predict(mdl_iR_vs_aad, cf, type = "response"))
#perform prediction on same datasets - new variable containing the prediction in log odds
cf <- cf %>%
  mutate(lo_iR_vs_aad = predict(mdl_iR_vs_aad, cf))
bpr <- predict(mdl_iR_vs_aad, cf, type = "response")
#visualisation
#how to I include the abline?
ggplot(cf, aes(AgeatDiagnosis, insulinRequire)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 20) +
  geom_smooth(method="glm", method.args = list(family = "binomial"))
#plotting - gives same plot essentially as prediction plot with ggplot above
effect_plot(mdl_iR_vs_aad, pred = AgeatDiagnosis, plot.points = TRUE,
            jitter = c(0.1, 0.05), point.alpha = 0.1) +
  ylab("Pr(insulinRequire = 1)")
```

##### 2.2.1.1 Binned vs log odds 
```{r bin aad}
#data bining plot
#identify decile cut-offs
quantile(cf$AgeatDiagnosis, probs = seq(0,1,0.1))
#set up cutoff values
wbreaks <- c(18, 29, 35, 38, 41, 43, 45, 47, 48, 50, 50)
#specify interval/bin labels
wtags <- c("[18, 29)", "[29, 35)", "[35, 38)", "[38, 41)", "[41,43)", "[43, 45)", "[45, 47)", "[47, 48)", "[48, 50)", "[50, 60)")
#sort values into decile bins
w <- cf %>%
  dplyr::select(AgeatDiagnosis, insulinRequire) #pick the variable 
wgroup <- as_tibble(w) %>% 
  mutate(tag = case_when(
    AgeatDiagnosis >= 18 & AgeatDiagnosis < 29 ~ wtags[1],
    AgeatDiagnosis >= 29 & AgeatDiagnosis < 35 ~ wtags[2],
    AgeatDiagnosis >= 35 & AgeatDiagnosis < 38 ~ wtags[3],
    AgeatDiagnosis >= 38 & AgeatDiagnosis < 41 ~ wtags[4],
    AgeatDiagnosis >= 41 & AgeatDiagnosis < 43 ~ wtags[5],
    AgeatDiagnosis >= 43 & AgeatDiagnosis < 45 ~ wtags[6],
    AgeatDiagnosis >= 45 & AgeatDiagnosis < 47 ~ wtags[7],
    AgeatDiagnosis >= 47 & AgeatDiagnosis < 48 ~ wtags[8],
    AgeatDiagnosis >= 48 & AgeatDiagnosis < 50 ~ wtags[9],
    AgeatDiagnosis > 50 ~ wtags[10],
    ))
summary(wgroup)
wgroup <- wgroup %>%
  mutate(lo_iR_vs_aad = predict(mdl_iR_vs_aad, cf))
w_bin_log <- wgroup %>%
  group_by(tag) %>%
  summarise(median_score=median(lo_iR_vs_aad))
ggplot(w_bin_log, aes(tag, median_score)) +
  geom_point() +
  labs(x = 'Age at Diagnosis decile bins', y = 'median log odds of T1D')

```

#### 2.2.2 Transformations

#### 2.2.3 Spline transformation
```{r aad cubic splines}
#plot function requires variables to be in numeric format
#run with 3, 4 & 5 knots
#how do I get the summary stats to automatically be attached - error message when run in chunk
#rcspline.plot(as.numeric(cf$AgeatDiagnosis), as.numeric(cf$insulinRequire), model="logistic", nk=3)
#rcspline.plot(as.numeric(cf$AgeatDiagnosis), as.numeric(cf$insulinRequire), model="logistic", nk=4)
#rcspline.plot(as.numeric(cf$AgeatDiagnosis), as.numeric(cf$insulinRequire), model="logistic", nk=5)

#run a non spline log regression, then one each with knots = 3, 4, 5 respectively
m0 <- lrm(insulinRequire ~ AgeatDiagnosis, data = cf)
m1 <- lrm(insulinRequire ~ rcs(AgeatDiagnosis, 3), data = cf)
m2 <- lrm(insulinRequire ~ rcs(AgeatDiagnosis, 4), data = cf)
m3 <- lrm(insulinRequire ~ rcs(AgeatDiagnosis, 5), data = cf)

dd <- datadist(cf)
options(datadist="dd")
summary(m0)
summary(m1)
summary(m2)
summary(m3)

#run anovas on each to see Wald statistics
anova(m0)
anova(m1)
anova(m2)
anova(m3)

#get aic
#why is this different to the plot AIC values for m1 etc.
mAIC <- c(m0 = AIC(m0), m1 = AIC(m1), m2 = AIC(m2), m3 = AIC(m3))
mAIC

```
*no knots*
AIC = 787.3316

*knots = 3*
n=1352, d=179
x2 = 278.08
AIC = 274.08
Wald association 223.22, p=0.0000
Wald linearity 4.511, p=0.0337* (anova)

*knots = 4*
n=1352, d=179
x2 = 278.78
AIC = 272.78
Wald association 221.02, p=0.0000
Wald linearity 5.04, p=0.0803 (anova)

*knots = 5*
n=1352, d=179
x2 = 279.00
AIC = 271.00
Wald association 219.53, p=0.0000
Wald linearity 5.23, p=0.1557 (anova)

PLOT
Relationship seems linear, knots = 5 seems best according to AIC, Wald association is 0.000 (stat sig for all three), but only stat sig for knots = 3
AIC TBL
Agrees with Wald association - lowest is for 3 knots

### Calibration
```{r calibration}
###extract quantiles for predicted probabilities - how do I get this to focus on log odds and then plot this against the quantiles?
bmi_valprob <- val.prob(p = pr, y = cf$insulinRequire, logit = cf$lo_iR_vs_bmi, group = cf$bmi, g.group = 10)
plot(bmi_valprob, xlab = "predicted probability", ylab = "actual probability")
```
